\artus -- Framework for Event-based Data Analysis in High Energy Physics

### 2015-09-23 Core/doc/latex/paper.tex
paper.tex: improved placement of figures
paper.tex: make todos inline to better read them

\begin{abstract}
We present \artus, an event-based data-processing framework for high energy physics experiments.
\mbox{\artus} is designed for large-scale data analysis in a collaborative environment.
The architecture design choices take into account typical challenges and are based on experiences with similar applications.
The structure of the framework and its advantages are described.
An example use case and performance measurements are presented.
The framework is well-tested and successfully used by several analysis groups.\todo{Mention them, maybe with publications? -> mentioned at the beginning of Htt chapter. Do we have publications?}
\end{abstract}

@@

The reconstruction of the raw data coming from the experiments is performed on large computing centres organized into the WLCG\footnote{\url{http://wlcg-public.web.cern.ch}}, the Worldwide LHC Computing Grid. The resulting datasets contain event information and high-level physics objects and are of $\mathcal O(10-100)$~TB in size. \todo{move file size numbers to Figure \ref{figure_analysis_steps}?}

@@

\begin{figure}[p]
\centering \input{figures/analysis_steps}
\caption[Characteristics of an event-based analysis in high energy physics]{Characteristic of an event-based analysis in high energy physics, numbers taken from \cite{joram_phd}. Data size is reduced by six orders of magnitude, turnaround time by a factor of 4. I/O intensive operations are done in highly parallel in the WLCG.
The data reduction is achieved by performing the data-intensive operations only rarely and applying sensible event selection steps as early as possible.}
\label{figure_analysis_steps}
\end{figure}
\todo{Should we drop CMS-specific technical terms like AOD/miniAOD?
Maybe not put this figure on a dedicated page?}
\clearpage

@@

Following the structure of the CMSSW \todo{Referenz? zumindest github-link?} framework, the \artus framework\footnote{\url{https://github.com/artus-analysis/\artus}} provides a front-to-end solution for the analysis of event-based data.
The basic concept of modules processing events within an event loop is complemented by tools for the configuration of every step.
The complete structure is explained in the next section~\ref{section_artus_structure}.

@@

A pipeline consists of an arbitrary number of processor modules that are run in a sequence.
Three kinds of processors are distinguished:
\begin{description}
\item[Producers] are meant to calculate new quantities based on the information in the event and the product and following the settings of the pipeline.
\item[Filters] apply a selection based on event properties.
In case an event does not fulfil a requirement defined by a filter, the processing of the subsequent producers and filters can be skipped.
\item[Consumers] can be added to local pipelines.
They are executed after the producers and filters.
The task of consumers is to take quantities from the event and the product and store them in the output according to the pipeline settings.
\end{description}
Producers and filters can be configured in an arbitrary order.
Consumers are executed after the chain of producers and filters.
The output together with the complete configuration is written to a ROOT \todo{Referenz auf ROOT?} file.
These processors are created and managed by a dedicated factory according to the configuration.

\todo{I would not put the remaining part of this chapter here.
It should be either part of the requirements or part of the summary what Artus does achieve; but in this technical part in between it feels misplaced
-> Should we put this part into a dedicated section directly after this one?
I think the idea was to explain how certain design decisions, e.g. pipelines, achieve the \artus design goals.
Therefore it shouldnt be too far from the technical descriptions.
}
This structure extends the established concepts of existing frameworks like FWLite~\cite{FWLite} from CMSSW to the following four main advantages:
\begin{description}
\item[Modular structure] Analysis steps are implemented in processors, breaking down the complete analysis in well-defined and easily understandable and writeable parts.
\item[Re-utilisation of code] Processors can be shared among different analysis groups.
Every user contributes to the testing of existing code and improves the confidence in this code.
\item[Configurability] Each module uses settings that are read from the the configuration.
Each pipeline is mapped to a separate sub-set of the configuration.
\item[High performance] The concept of pipelines allows to analyse the data with different configurations at the same time, while the data is only read in only once and common processing steps are shared in the global pipeline.
Therefore, the cost-intensive input operations are reduced to a minimum.
\end{description}

\section{Building an End-user Analysis \label{section_artus_analysis}}
\todo{If I wouldn't know the structure already, the following part wouldn't have helped my.
Either we support it with a drawing and more detailled explanations or we just refer to a readme...
Maybe rewrite parts of this section}
In the \artus framework, each concrete analysis is defined by a set of object types.
The analysis programmer must provide an implementation of the classes for settings, the event and the product class.
The settings class defines the tags that are read from the configuration file and the corresponding types of these values.
The event class defines the event content which is read from the input and the product class contains members for new quantities calculated during the analysis stage.
These three classes need to be derived from the base classes provided by the core of the \artus framework.

@@

The structure of the framework is illustrated in figure~\ref{figure_artus_harry_plotter}.
It follows the idea of processors in the \artus framework.
Three kinds of processor modules are executed subsequently:
\begin{enumerate}
\item One or, in special cases, more input modules are run to read in data from an input source.
The input format is defined by the input module.
The most commonly used input module is the one that reads histograms or other objects from ROOT files.
\item An arbitrary number of analysis modules is appended.
They can perform calculations based on the inputs or the results of previously run analysis modules.
\item One or, in special cases, multiple plot modules are executed in the last step.
Their task is to output the processed data in the specified format.
In most cases, the output is a plot but it is also possible to store the objects in new ROOT files.
\end{enumerate}
All processors share a common plot data object and possible meta-data where they can add new information and read the existing one.
For example, the input modules add histograms read from input files and the plot modules access these histograms.

>>>>

\begin{abstract}
We present \artus, an event-based data-processing framework for high energy physics experiments.
\mbox{\artus} is designed for large-scale data analysis in a collaborative environment.
The architecture design choices take into account typical challenges and are based on experiences with similar applications.
The structure of the framework and its advantages are described.
An example use case and performance measurements are presented.
The framework is well-tested and successfully used by several analysis groups.\todo[inline]{Mention them, maybe with publications? -> mentioned at the beginning of Htt chapter. Do we have publications?}
\end{abstract}

@@

The reconstruction of the raw data coming from the experiments is performed on large computing centres organized into the WLCG\footnote{\url{http://wlcg-public.web.cern.ch}}, the Worldwide LHC Computing Grid. The resulting datasets contain event information and high-level physics objects and are of $\mathcal O(10-100)$~TB in size. \todo[inline]{move file size numbers to Figure \ref{figure_analysis_steps}?}

@@

\begin{figure}[t]
\centering \input{figures/analysis_steps}
\caption[Characteristics of an event-based analysis in high energy physics]{Characteristic of an event-based analysis in high energy physics, numbers taken from \cite{joram_phd}. Data size is reduced by six orders of magnitude, turnaround time by a factor of 4. I/O intensive operations are done in highly parallel in the WLCG.
The data reduction is achieved by performing the data-intensive operations only rarely and applying sensible event selection steps as early as possible.}
\label{figure_analysis_steps}
\end{figure}

\todo[inline]{Should we drop CMS-specific technical terms like AOD/miniAOD?
Maybe not put this figure on a dedicated page?}

@@

Following the structure of the CMSSW \todo[inline]{Referenz? zumindest github-link?} framework, the \artus framework\footnote{\url{https://github.com/artus-analysis/\artus}} provides a front-to-end solution for the analysis of event-based data.
The basic concept of modules processing events within an event loop is complemented by tools for the configuration of every step.
The complete structure is explained in the next section~\ref{section_artus_structure}.

@@

A pipeline consists of an arbitrary number of processor modules that are run in a sequence.
Three kinds of processors are distinguished:
\begin{description}
\item[Producers] are meant to calculate new quantities based on the information in the event and the product and following the settings of the pipeline.
\item[Filters] apply a selection based on event properties.
In case an event does not fulfil a requirement defined by a filter, the processing of the subsequent producers and filters can be skipped.
\item[Consumers] can be added to local pipelines.
They are executed after the producers and filters.
The task of consumers is to take quantities from the event and the product and store them in the output according to the pipeline settings.
\end{description}
Producers and filters can be configured in an arbitrary order.
Consumers are executed after the chain of producers and filters.
The output together with the complete configuration is written to a ROOT \todo[inline]{Referenz auf ROOT?} file.
These processors are created and managed by a dedicated factory according to the configuration.

\todo[inline]{I would not put the remaining part of this chapter here.
It should be either part of the requirements or part of the summary what Artus does achieve; but in this technical part in between it feels misplaced
-> Should we put this part into a dedicated section directly after this one?
I think the idea was to explain how certain design decisions, e.g. pipelines, achieve the \artus design goals.
Therefore it shouldnt be too far from the technical descriptions.
}
This structure extends the established concepts of existing frameworks like FWLite~\cite{FWLite} from CMSSW to the following four main advantages:
\begin{description}
\item[Modular structure] Analysis steps are implemented in processors, breaking down the complete analysis in well-defined and easily understandable and writeable parts.
\item[Re-utilisation of code] Processors can be shared among different analysis groups.
Every user contributes to the testing of existing code and improves the confidence in this code.
\item[Configurability] Each module uses settings that are read from the the configuration.
Each pipeline is mapped to a separate sub-set of the configuration.
\item[High performance] The concept of pipelines allows to analyse the data with different configurations at the same time, while the data is only read in only once and common processing steps are shared in the global pipeline.
Therefore, the cost-intensive input operations are reduced to a minimum.
\end{description}

\section{Building an End-user Analysis \label{section_artus_analysis}}
\todo[inline]{If I wouldn't know the structure already, the following part wouldn't have helped my.
Either we support it with a drawing and more detailled explanations or we just refer to a readme...
Maybe rewrite parts of this section}
In the \artus framework, each concrete analysis is defined by a set of object types.
The analysis programmer must provide an implementation of the classes for settings, the event and the product class.
The settings class defines the tags that are read from the configuration file and the corresponding types of these values.
The event class defines the event content which is read from the input and the product class contains members for new quantities calculated during the analysis stage.
These three classes need to be derived from the base classes provided by the core of the \artus framework.

@@

The structure of the framework is illustrated in Figure~\ref{figure_artus_harry_plotter}.
It follows the idea of processors in the \artus framework.
Three kinds of processor modules are executed subsequently:
\begin{enumerate}
\item One or, in special cases, more input modules are run to read in data from an input source.
The input format is defined by the input module.
The most commonly used input module is the one that reads histograms or other objects from ROOT files.
\item An arbitrary number of analysis modules is appended.
They can perform calculations based on the inputs or the results of previously run analysis modules.
\item One or, in special cases, multiple plot modules are executed in the last step.
Their task is to output the processed data in the specified format.
In most cases, the output is a plot but it is also possible to store the objects in new ROOT files.
\end{enumerate}
All processors share a common plot data object and possible meta-data where they can add new information and read the existing one.
For example, the input modules add histograms read from input files and the plot modules access these histograms.
